God in Coq: \\ Modal Logic in Coq \\ and Scott's Version of G\"{o}del's Proof of God's Existence \\ {\small (\emph{Rough Diamond Proof Pearl})}

### 2014-01-30 Papers/ITP/GodProof-ITP.tex
update of ITP paper

\author{
  Christoph Benzm\"{u}ller\inst{1} 
  \and 
  Bruno Woltzenlogel Paleo\inst{2}
}

@@

A crucial aspect of modal logics \cite{ToDo} is that the so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$ is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive attempts to define the modal operators $\Box$ and $\Diamond$ may easily be unsound in this respect. To avoid this issue, the \emph{possible world semantics} of modal logics can be explicitly embedded into higher-order logics \cite{J23,B9}. To this aim, a type for worlds must be declared and modal propositions should be not of type \texttt{Prop} but of a lifted type that depends on possible worlds. Possible worlds are connected by an accessibility relation.

@@

\begin{small}
\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Parameter i: \Type. \com{(* Type for worlds *)}
\Parameter u: \Type. \com{(* Type for individuals *)}
\Definition o := i -> \Prop. \com{(* Type of modal propositions *)}
\Parameter r: i -> i -> \Prop. \com{(* Acessibility relation for worlds *)}
\end{Verbatim}
\end{small}

@@

\noindent
Although the embedding via the possible world semantics is necessary, it should be as transparent as possible to the user. Fortunately, basic Coq tactics such \texttt{intro}, \texttt{apply} and \texttt{split} automatically unfold the modal definitions and thus can be used with modal connectives and quantifiers as they are used with the usual connectives and quantifiers. Nothing else needs to be done. For the modal operators, however, the following simple tactics have been implemented, in order to allow the user to work with the concepts of \emph{necessity} and \emph{possibility} without having to unfold the definitions of modal operators and think in terms of possible worlds.

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Theorem dia_box_to_box: [ mforall p, (dia (box p)) m-> (box p) ].
\Proof. mv. intro p. intro H1. dia_e H1. box_i. box_e H1 H2. exact H2. eapply transitivity.
  apply symmetry. exact R.

  exact R0.
\Qed.
\end{Verbatim}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  
G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}). The formalization shown in this section aims at
being as similar as possible to Dana Scott's version of the proof \cite{ScottNotes}. The formulation and numbering of axioms, definitions and theorems is the same as in Scott's notes. Even the Coq proof scripts follow all the steps in Scott's notes. Scott's assertions are emphasized with comments. In contrast to the formalization in Isabelle \cite{ToDo:our isabelle formalization in ArchiveFormalProofs}, where automation via Metis \cite{Hurd03first-orderproof} and Sledgehammer \cite{Sledgehammer} using tools such as Nitpick \cite{Nitpick}, LEO-II \cite{LEO-II} and Satallax \cite{Satallax} has been successfully employed, the formalization in Coq used no automation. This was a deliberate choice, mainly because it allowed a qualitative evaluation of the convenience of the possible world semantic embedding approach for \emph{interactive} theorem proving. Moreover, in order to formalize precisely Scott's version and not another automatically found version, automation would have to be heavily limited anyway. Furthermore, the deliberate preference for simple tactics (mostly \emph{intro}, \emph{apply} and the modal tactics described in the previous section) results in proof scripts that could be read as natural deduction proofs. This hopefully makes the formalization more accessible to those who are not experts in Coq's tactics but are nevertheless interested in G\"odel's proof.

@@

\com{(* Theorem T1: positive properties are possibly exemplified *)}
\Theorem theorem1: [ mforall p, (Positive p) m-> dia (mexists x, p x) ].
\Proof. mv.
intro p. intro H1. proof_by_contradiction H2. apply not_dia_box_not in H2.
assert (H3: ((box (mforall x, m~ (p x))) w)). \com{(* Scott *)}
  box_i. intro x. assert (H4: ((m~ (mexists x : u, p x)) w0)).
    box_e H2 G2. exact G2.

    clear H2 R H1 w. intro H5. apply H4. exists x. exact H5.

  assert (H6: ((box (mforall x, (p x) m-> m~ (x m= x))) w)). \com{(* Scott *)}   
    box_i. intro x. intros H7 H8. box_elim H3 w0 G3. eapply G3. exact H7.

    assert (H9: ((Positive (fun x => m~ (x m= x))) w)). \com{(* Scott *)}
      apply (axiom2 w p (fun x => m~ (x m= x))). split.
        exact H1.

        exact H6.
      assert (H10: ((box (mforall x, (p x) m-> (x m= x))) w)). \com{(* Scott *)}
        box_i. intros x H11. reflexivity.

        assert (H11 : ((Positive (fun x => (x m= x))) w)). \com{(* Scott *)}
          apply (axiom2 w p (fun x => x m= x )). split.
            exact H1.

            exact H10.

          apply axiom1a in H9. contradiction.
\Qed.

@@

\com{(* Theorem T2: being God-like is an essence of any God-like being *)}
\Theorem theorem2: [ mforall x, (G x) m-> (G ess x) ].
\Proof. mv. intro g. intro H1. unfold Essence. split.
  exact H1.

  intro q. intro H2. assert (H3: ((Positive q) w)).
    proof_by_contradiction H4. unfold G in H1. apply axiom1b in H4. apply H1 in H4. 
    contradiction.

    cut (box (Positive q) w). \com{(* Scott *)}
      apply K. box_i. intro H5. intro y. intro H6. unfold G in H6. apply (H6 q). exact H5.

      apply axiom4. exact H3.
\Qed.

@@

\Lemma lemma1: [ (mexists z, (G z)) m-> box (mexists x, (G x)) ].
\Proof. mv. intro H1. destruct H1 as [g H2]. cut ((G ess g) w). \com{(* Scott *)}
  assert (H3: (NE g w)).       \com{(* Scott *)}
    unfold G in H2. apply (H2 NE). apply axiom5.

    unfold NE in H3. apply H3.

  apply theorem2. exact H2.
\Qed.

\Lemma lemma2: [ dia (mexists z, (G z)) m-> box (mexists x, (G x)) ].
\Proof. mv. intro H. cut (dia (box (mexists x, G x)) w).  \com{(* Scott *)}
  apply dia_box_to_box.

  apply (mp_dia w (mexists z, G z)).
    exact H.

    box_i. apply lemma1.
\Qed.

@@

The theological implications of the verified correctness of 
this proof certainly depend on a critical discussion of the 
underlying concepts, definitions and axioms, 
which is beyond the scope of this paper. Nevertheless, the application
of theorem proving technology to a philosophical domain 
has been fruitful for both areas. New results of philosophical 
interest have been obtained \cite{ToDo:our ArXiv or Journal abstract} 
and basic infra-structure for interactive and automated reasoning 
in higher-order modal logics has been created.

\marginpar{cite John Rushby's formalization of anselm's argument?}

@@

\bibitem{Adams}
R.M. Adams.
\newblock Introductory note to *1970.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
  Letters}}. Oxford University Press, 1995.

\bibitem{AndersonGettings}
A.C. Anderson and M.~Gettings.
\newblock G\"odel ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}. {Springer}, 1996.

\bibitem{B9}
C.~Benzm{\"u}ller and L.C. Paulson.
\newblock Exploring properties of normal multimodal logics in simple type
  theory with {LEO-II}.
\newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
  Birthday}}, pages 386--406. College Publications.

\bibitem{J23}
C.~Benzm{\"u}ller and L.C. Paulson.
\newblock Quantified multimodal logics in simple type theory.
\newblock {\em Logica Universalis (Special Issue on Multimodal Logics)},
  7(1):7--20, 2013.

@@

@@

\bibitem{Sledgehammer}
J.C. Blanchette, S.~B\"ohme, and L.C. Paulson.
\newblock Extending {Sledgehammer} with {SMT} solvers.
\newblock {\em Journal of Automated Reasoning}, 51(1):109--128, 2013.

\bibitem{Nitpick}
J.C. Blanchette and T.~Nipkow.
\newblock Nitpick: A counterexample generator for higher-order logic based on a
  relational model finder.
\newblock In {\em Proc. of ITP 2010}, number 6172 in LNCS, pages 131--146.
  Springer, 2010.

\bibitem{Satallax}
C.E. Brown.
\newblock Satallax: An automated higher-order prover.
\newblock In {\em Proc. of IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117.
  Springer, 2012.

@@

@@

\bibitem{Hazen}
A.P. Hazen.
\newblock On g\"odel's ontological proof.
\newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

@@

\bibitem{FormalTheologyRepository}
B.~Woltzenlogel Paleo and C.~Benzm\"uller.
\newblock Formal theology repository
  (\url{https://github.com/FormalTheology/GoedelGod}).

@@

\bibitem{sobel2004logic}
J.H. Sobel.
\newblock {\em Logic and Theism: Arguments for and Against Beliefs in God}.
\newblock Cambridge U. Press, 2004.

@@

\end{thebibliography}

>>>>

\author{
  Christoph Benzm\"{u}ller\inst{1}\thanks{This work has been supported by the German Research Foundation (DFG) under grant BE2501/9-1.} 
  \and 
  Bruno Woltzenlogel Paleo\inst{2}
}

@@

A crucial aspect of modal logics \cite{ModalLogic} is that the so-called \emph{necessitation rule} allows $\Box A$ to be derived if $A$ is a theorem, but $A \imp \Box A$ is not necessarily a theorem. Naive attempts to define the modal operators $\Box$ and $\Diamond$ may easily be unsound in this respect. To avoid this issue, the \emph{possible world semantics} of modal logics can be explicitly embedded into higher-order logics \cite{J23,B9}. To this aim, a type for worlds must be declared and modal propositions should be not of type \texttt{Prop} but of a lifted type that depends on possible worlds. Possible worlds are connected by an accessibility relation.

The embedding is related to labeling techniques
\cite{Labels}. However, the expressiveness of higher-order logic is
exploited here to encode the labels within the logic language itself.

@@

\begin{small}
\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Parameter i: \Type. \com{(* Type for worlds *)}
\Parameter u: \Type. \com{(* Type for individuals *)}
\Definition o := i -> \Prop. \com{(* Type of modal propositions *)}
\Parameter r: i -> i -> \Prop. \com{(* Accessibility relation for worlds *)}
\end{Verbatim}
\end{small}

@@

\noindent
To maximally preserve user intuition in interactive modal logic theorem
proving, the embedding via the possible world semantics should be as
transparent as possible to the user. Fortunately, basic Coq tactics
such \texttt{intro}, \texttt{apply} and \texttt{split} automatically
unfold the modal definitions. They can thus be used with modal
connectives and quantifiers just as they are used with the usual
connectives and quantifiers.  For the modal operators, however, the
following simple tactics have been implemented, in order to allow the
user to work with the concepts of \emph{necessity} and
\emph{possibility} without having to unfold the definitions of modal
operators and think in terms of possible worlds.

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Theorem dia_box_to_box: [ mforall p, (dia (box p)) m-> (box p) ].
\Proof. mv. intro p. intro H1. dia_e H1. box_i. box_e H1 H2. exact H2. eapply transitivity.
  apply symmetry. exact R.
  exact R0.
\Qed.
\end{Verbatim}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  
G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}). The formalization shown in this section aims at
being as similar as possible to Dana Scott's version of the proof \cite{ScottNotes}. The formulation and numbering of axioms, definitions and theorems is the same as in Scott's notes. Even the Coq proof scripts follow all the steps in Scott's notes. Scott's assertions are emphasized with comments. In contrast to the formalization in Isabelle \cite{Archiv}, where automation via Metis \cite{Hurd03first-orderproof} and Sledgehammer \cite{Sledgehammer} using tools such as Nitpick \cite{Nitpick}, LEO-II \cite{LEO-II} and Satallax \cite{Satallax} has been successfully employed, the formalization in Coq used no automation. This was a deliberate choice, mainly because it allowed a qualitative evaluation of the convenience of the possible world semantic embedding approach for \emph{interactive} theorem proving. Moreover, in order to formalize precisely Scott's version and not another automatically found version, automation would have to be heavily limited anyway. Furthermore, the deliberate preference for simple tactics (mostly \emph{intro}, \emph{apply} and the modal tactics described in the previous section) results in proof scripts that could be read as natural deduction proofs. This hopefully makes the formalization more accessible to those who are not experts in Coq's tactics but are nevertheless interested in G\"odel's proof.

@@

\com{(* Theorem T1: positive properties are possibly exemplified *)}
\Theorem theorem1: [ mforall p, (Positive p) m-> dia (mexists x, p x) ].
\Proof. mv.
intro p. intro H1. proof_by_contradiction H2. apply not_dia_box_not in H2.
assert (H3: ((box (mforall x, m~ (p x))) w)). \com{(* Scott *)}
  box_i. intro x. assert (H4: ((m~ (mexists x : u, p x)) w0)).
    box_e H2 G2. exact G2.
    clear H2 R H1 w. intro H5. apply H4. exists x. exact H5.
  assert (H6: ((box (mforall x, (p x) m-> m~ (x m= x))) w)). \com{(* Scott *)}   
    box_i. intro x. intros H7 H8. box_elim H3 w0 G3. eapply G3. exact H7.
    assert (H9: ((Positive (fun x => m~ (x m= x))) w)). \com{(* Scott *)}
      apply (axiom2 w p (fun x => m~ (x m= x))). split.
        exact H1.
        exact H6.
      assert (H10: ((box (mforall x, (p x) m-> (x m= x))) w)). \com{(* Scott *)}
        box_i. intros x H11. reflexivity.
        assert (H11 : ((Positive (fun x => (x m= x))) w)). \com{(* Scott *)}
          apply (axiom2 w p (fun x => x m= x )). split.
            exact H1.
            exact H10.
          apply axiom1a in H9. contradiction.
\Qed.

@@

\com{(* Theorem T2: being God-like is an essence of any God-like being *)}
\Theorem theorem2: [ mforall x, (G x) m-> (G ess x) ].
\Proof. mv. intro g. intro H1. unfold Essence. split.
  exact H1.
  intro q. intro H2. assert (H3: ((Positive q) w)).
    proof_by_contradiction H4. unfold G in H1. apply axiom1b in H4. apply H1 in H4. 
    contradiction. 
    cut (box (Positive q) w). \com{(* Scott *)}
      apply K. box_i. intro H5. intro y. intro H6. unfold G in H6. apply (H6 q). exact H5.
      apply axiom4. exact H3.
\Qed.

@@

\Lemma lemma1: [ (mexists z, (G z)) m-> box (mexists x, (G x)) ].
\Proof. mv. intro H1. destruct H1 as [g H2]. cut ((G ess g) w). \com{(* Scott *)}
  assert (H3: (NE g w)).       \com{(* Scott *)}
    unfold G in H2. apply (H2 NE). apply axiom5.
    unfold NE in H3. apply H3.
  apply theorem2. exact H2.
\Qed.

\Lemma lemma2: [ dia (mexists z, (G z)) m-> box (mexists x, (G x)) ].
\Proof. mv. intro H. cut (dia (box (mexists x, G x)) w).  \com{(* Scott *)}
  apply dia_box_to_box.
  apply (mp_dia w (mexists z, G z)).
    exact H.
    box_i. apply lemma1.
\Qed.

@@

The theological implications of the verified correctness of this proof
certainly depend on a critical discussion of the underlying concepts,
definitions and axioms, which is beyond the scope of this paper. What
can be said though is that our work presents further evidence that
theism is at least not (obviously) irrational.

Related work includes formalizations of Anselm's argument by Rushby
\cite{Rushby}, and by Oppenheimer and Zalta \cite{Zalta}.  Obviously,
the application of theorem proving technology to the domain of
theoretical philosophy can --- as already pictured by Leibniz ---
be very fruitful for both areas. In fact, new results of
philosophical interest have been obtained by our
work~\cite{arXiv}. Moreover a first, basic infrastructure for
interactive and automated reasoning in higher-order modal logics has
been created and will be further improved in the future.

@@

\bibitem{Adams}
R.M.~Adams.
\newblock Introductory note to *1970.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
  Letters}}. Oxford University Press, 1995.

\bibitem{AndersonGettings}
A.C.~Anderson and M.~Gettings.
\newblock G\"odel ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}. {Springer}, 1996.

\bibitem{B9}
C.~Benzm{\"u}ller and L.C.~Paulson.
\newblock Exploring properties of normal multimodal logics in simple type
  theory with {LEO-II}.
\newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
  Birthday}}, pages 386--406. College Publications.

\bibitem{J23}
C.~Benzm{\"u}ller and L.C.~Paulson.
\newblock Quantified multimodal logics in simple type theory.
\newblock {\em Logica Universalis (Special Issue on Multimodal Logics)},
  7(1):7--20, 2013.

@@

\bibitem{arXiv}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock Formalization, Mechanization and Automation of G{\"o}del's Proof of God's Existence.
\newblock {\em {arXiv}}:1308.4526, 2013.

\bibitem{arXiv}
C.~Benzm{\"u}ller and B.~Woltzenlogel Paleo,
\newblock G{\"o}del's God in Isabelle/HOL. 
\newblock {\em Archive of Formal Proofs}, 2013.

@@

\bibitem{ModalLogic}
P.~Blackburn, J.v.~Benthem, and F.~Wolter (eds.), \newblock {\em {Handbook of Modal Logic}}. Elsevier, 2006.

\bibitem{Sledgehammer}
J.C.~Blanchette, S.~B\"ohme, and L.C.~Paulson.
\newblock Extending {Sledgehammer} with {SMT} solvers.
\newblock {\em Journal of Automated Reasoning}, 51(1):109--128, 2013.

\bibitem{Nitpick}
J.C.~Blanchette and T.~Nipkow.
\newblock Nitpick: A counterexample generator for higher-order logic based on a
  relational model finder.
\newblock In {\em Proc. of ITP 2010}, no.~6172 in LNCS, pages 131--146.
  Springer, 2010.

\bibitem{Zalta}
P.E.~Oppenheimera and E.N.~Zalta.
\newblock A Computationally-Discovered Simplification of the Ontological Argument.
\newblock {\em Australasian Journal of Philosophy}, 89(2):333-349, 2011.

\bibitem{Satallax}
C.E.~Brown.
\newblock Satallax: An automated higher-order prover.
\newblock In {\em Proc. of IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117.
  Springer, 2012.

@@

\bibitem{Labels}
D.M.~Gabbay. {\em Labelled Deductive Systems}. Clarendon Press, 1996.

@@

\bibitem{Hazen}
A.P.~Hazen.
\newblock On g\"odel's ontological proof.
\newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

@@

\bibitem{Rushby}
J.~Rushby.
\newblock The Ontological Argument in PVS. 
\newblock In {\em Proc.~of CAV Workshop ``Fun With Formal Methods''}, St. Petersburg, Russia, 2013.

@@

\bibitem{sobel2004logic}
J.H.~Sobel.
\newblock {\em Logic and Theism: Arguments for and Against Beliefs in God}.
\newblock Cambridge U. Press, 2004.

@@

\bibitem{FormalTheologyRepository}
B.~Woltzenlogel Paleo and C.~Benzm\"uller.
\newblock Formal theology repository
  (\url{https://github.com/FormalTheology/GoedelGod}).
\end{thebibliography}
